---
title: "Sicily Sailing GPS track summary"
author: jsp
date: today
format: 
  html:
    code-fold: true
    toc: true
editor: visual
editor_options: 
  chunk_output_type: console
---

```{r}
#| include: false
library(XML)
library(lubridate)
library(celestial)
library(leaflet)
library(tidyverse)
library(traipse)

theme_set(theme_bw())
```

Navionics GPs track of the Sicily eolian island trip.

::: panel-tabset
## Hide {-}
## Track loading and data formatting {-}

The xml file with the navionics tracks is parsed.

-   Coordinates, time and speed are extracted.
-   Speed is converted from m/s to knots
-   Distance is calculated from thinned 1min segment data for the total distance and cumulative per day
-   COG is calculated as consecutive bearings from a thinned 10min segment dataset

```{r}
# temp <- tempfile(fileext = ".zip")
# download.file("https://drive.google.com/drive/folders/1FlAM74hAPtFC3qIcZF5nQ03UK-WP3hZc&export=download",
#   temp)
# out <- unzip(temp, exdir = tempdir())
# gpx_parsed <- htmlTreeParse(file = out[14], useInternalNodes = TRUE)

gpx_parsed <- htmlTreeParse(file = "Sicily2021Tracks", useInternalNodes = TRUE)

coords <- xpathSApply(doc = gpx_parsed, path = "//trkpt", fun = xmlAttrs)
date_time <- xpathSApply(doc = gpx_parsed, path = "//trkpt/time", fun = xmlValue)
speed <- xpathSApply(doc = gpx_parsed, path = "//trkpt/extensions/navionics_speed", fun = xmlValue)
elevation <- xpathSApply(doc = gpx_parsed, path = "//trkpt/ele", fun = xmlValue)

smoothing_window <- "30 sec"

df <- data.frame(
  lat = as.numeric(coords["lat", ]),
  lon = as.numeric(coords["lon", ]),
  date = lubridate::as_date(date_time),
  time = lubridate::parse_date_time(date_time, "Y-m-d*H:M:S") + hours(2), #switch timezone
  kts = as.numeric(speed) * 1.944,
  elevation = as.numeric(elevation)) |> # convert from m/s to knots 
  mutate(lat_smooth = runner::runner(lat, k = smoothing_window, idx = time,
                                   f = function(x) median(x, na.rm = TRUE)),
         lon_smooth = runner::runner(lon, k = smoothing_window, idx = time,
                                   f = function(x) median(x, na.rm = TRUE))
         ) |>
  mutate(
    segments_full = track_distance(lat, lon) / 1852, # convert from meters to nm
    segments_full_smooth = track_distance(lat_smooth, lon_smooth) / 1852, # convert from meters to nm
  ) 
df$segments_full[1] <- 0
df$segments_full_smooth[1] <- 0
df <- df |>
  mutate(dist = cumsum(segments_full) |> round(digits = 1)) |>
  group_by(date) |>
  mutate(dist_day = cumsum(segments_full) |> round(digits = 1)) |> 
  ungroup() |>
  mutate(dist_sm = cumsum(segments_full_smooth) |> round(digits = 1)) |>
  group_by(date) |>
  mutate(dist_day_sm = cumsum(segments_full_smooth) |> round(digits = 1)) |> 
  ungroup()
# some success smoothing GPS points
# For each point: take all points received within 5 seconds of that point, calculate median of latitude and longitude individually, then use median values as latitude and longitude for that point. This generally gives a straight walk along a sidewalk. For driving, it clips corners, though.

# outlier check
outlier <- which(df$segments_full > .5)
outlier_area <- c(outlier, outlier-1) |> sort()

df_1min <- df |>
  dplyr::filter(second(time) < 1) |>
  mutate(cog = track_bearing(lat_smooth, lon_smooth),
         cog = ifelse(cog < 0, cog + 360, cog)
  ) |>
  mutate(
    segment = track_distance(lat, lon) / 1852, # convert from meters to nm
    segment_sm = track_distance(lat_smooth, lon_smooth) / 1852, # convert from meters to nm
  )
df_1min$segment[1] <- 0
df_1min$segment_sm[1] <- 0
df_1min <- df_1min |>
  mutate(dist = cumsum(segment) |> round(digits = 1)) |>
  group_by(date) |>
  mutate(dist_day = cumsum(segment) |> round(digits = 1)) |> 
  ungroup() |>
  mutate(dist_sm = cumsum(segment_sm) |> round(digits = 1)) |>
  group_by(date) |>
  mutate(dist_day_sm = cumsum(segment_sm) |> round(digits = 1)) |> 
  ungroup()

df_10min <- df_1min |>
  dplyr::filter(second(time) <1, minute(time) %% 10 == 0) |>
  mutate(cog = track_bearing(lat_smooth, lon_smooth),
         cog = ifelse(cog < 0, cog + 360, cog)) 

```

```{r, include=FALSE}
rm(coords, elevation, date_time, spees)
gc()
```

### example lines

```{r}
head(df, 5)
tail(df, 5)
```

Extract full hour data.

```{r}
df_1hour <- df_10min |> mutate(hour = hour(time)) |>
  group_by(date, hour) |>
  slice(1) |>
  ungroup() |> arrange(time)
  #dplyr::filter(minute(time) == 0, second(time) < 1) 
```

:::


# Trip overview

- Blue markers are the positions at the full hour for approximate logbook entries. 
- Red markers show beginning and end of missing segments.

::: panel-tabset
## Minute wise data

```{r minute_map}
#| fig-width: 5
#| fig-height: 5.5

leaflet() |>
  addTiles() |>
  addPolylines(data = df_1min, lat = ~lat, lng = ~lon, 
               color = "#000000", opacity = 0.8, weight = 3) |>
  leaflet::addAwesomeMarkers(
    data = df_1hour, lat = ~lat, lng = ~lon, label = ~as.character(hour(time))) |>
  leaflet::addAwesomeMarkers(
    data = df[outlier_area, ], lat = ~lat, lng = ~lon,
    label = ~as.character(hour(time)), 
    icon = awesomeIcons(icon = 'ios-close', iconColor = 'black',
                        library = 'ion', markerColor = 'red')) 
```

## Full data

```{r full_map}
#| fig-width: 5
#| fig-height: 5.5

leaflet() |>
  addTiles() |>
  addPolylines(data = df, lat = ~lat, lng = ~lon, 
               color = "#000000", opacity = 0.8, weight = 3) |>
  leaflet::addAwesomeMarkers(
    data = df_1hour, lat = ~lat, lng = ~lon, label = ~as.character(hour)) |>
  leaflet::addAwesomeMarkers(
    data = df[outlier_area, ], lat = ~lat, lng = ~lon,
    label = ~as.character(hour(time)), 
    icon = awesomeIcons(icon = 'ios-close', iconColor = 'black',
                        library = 'ion', markerColor = 'red'))
```


## Smoothed data

```{r full_map}
#| fig-width: 5
#| fig-height: 5.5

leaflet() |>
  addTiles() |>
  addPolylines(data = df, lat = ~lat_smooth, lng = ~lon, 
               color = "#000000", opacity = 0.8, weight = 3) |>
  leaflet::addAwesomeMarkers(
    data = df_1hour, lat = ~lat, lng = ~lon, label = ~as.character(hour)) |>
  leaflet::addAwesomeMarkers(
    data = df[outlier_area, ], lat = ~lat, lng = ~lon,
    label = ~as.character(hour(time)), 
    icon = awesomeIcons(icon = 'ios-close', iconColor = 'black',
                        library = 'ion', markerColor = 'red'))
```
:::

# Hourly log book

```{r logbook}
df_1hour |>
  mutate(latitude = deg2dms(lat, type = "cat"),
         longitude = deg2dms(lon, type = "cat"),
         # add distance
         across(c(cog, kts, dist), \(x) round(x, 1))
         ) |>
  mutate(time = sprintf("%02d:%02d", hour(time), minute(time)),
         date = str_remove(as.character(date), "^[0-9]*-"))|>
  select(date, time, latitude, longitude, cog, kts, dist_day, dist) |>
  gt::gt() |>
  gt::opt_interactive(use_text_wrapping = FALSE) |>
  gt::cols_width(dplyr::any_of(c("latitude", "longitude")) ~ gt::px(130))
```

# COG

::: panel-tabset
## 10 min thinning

```{r cog_10min}
#| warning: false
#| fig.width: 8
#| fig.height: 3

df_10min |>
  ggplot(aes(x = time, y = cog)) + geom_point() + geom_line()

```

## 1 min thinning

```{r cog_1min}
#| warning: false
#| fig.width: 8
#| fig.height: 3

df_1min |>
  ggplot(aes(x = time, y = cog)) + geom_point() + geom_line()
```
:::

# Distance

Checking Cumulative Miles in total and per day. The full track shows missing segments, marked in red in the full data route above.

::: panel-tabset
## 10 min thinning

```{r distance_10min}
#| warning: false
#| fig.width: 8
#| fig.height: 3

df_10min |>
  ggplot(aes(x = time, y = dist)) + geom_point() + geom_line()

df_10min |>
  ggplot(aes(x = time, y = dist_day)) + geom_point() + geom_line()
```

## full data

missing segments

```{r distance_full}
#| warning: false
#| fig.width: 8
#| fig.height: 3
df |>
  ggplot(aes(x = time, y = segments_full)) + geom_point() + geom_line()

df |>
  ggplot(aes(x = time, y = segments_full_smooth)) + geom_point() + geom_line()

```
:::
